\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-08-12 NAR_DRAFT.tex
check in changes

\subsection{Estimation of the mutation profile}
There are many aspects of genomic architecture which can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which has been shown to have a strong effect on the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context at a particular genomic site is made up of the identity of the reference base and the identity of the 3` and 5` flanking bases.
There are four possible bases {\{A,C,T,G\}} 3` and 5` of the site, leading to 16 possible flanking base combinations.
Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site {\{C,T\}}, resulting in $2 \cdot 16 = 32$ possible tri-nucleotide contexts.
We define a mutation context $C$ indexed by $i=\{1 \dots 32\}$, to be the tri-nucleotide context of a particular site in the genome. 
For each context, a mutation can be to any of the three possible alternate alleles {\{A,(C/T),G\}}. 
We can thus define $\mathrm{S}$, the substitution type, indexed by $i=\{1 \dots 32\}$ contexts and $j = \{1 \dots 3\}$ alternate bases, resulting in 96 possible substitution types.
Using the substitution type $\mathrm{S}$, we can rewrite equation \ref{eqn:full_prior} as

\begin{equation}
  \label{eqn:detailed_prior}
  $$
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S_{i,j}}) &=&  \mathrm{p}(m =j, M | C = i) \\
                            &=& \mathrm{p}(m = j \mid C = i) \mathrm{p}(C = i \mid M)\frac{p(M)}{p(C = i)},
  \end{array}
  $$
\end{equation}
where $\mathrm{p}(\mathrm{S_{i,j}})$ is the probability that a substitution of type $\mathrm{S_{i,j}}$ will occur.

@@

Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all observed tumor mutations of substitution type $\mathrm{S}_{i,j}$.
The distribution of observed substitutions $ \mathrm{p}(\mathrm{S}) $, is multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{i,j}\} $.
In a tumor with a very large number of verified mutations $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile, but in practice we find that many elements of $\boldsymbol{\pi}$ will have $0$ or a very small number of mutations present in the tumor.
As a result we need to model $\mathrm{p}(\mathrm{S})$ as dirichlet - multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$,

@@

In order to estimate $\boldsymbol{\pi}$, we generate a subset of variants in the tumor in which we have high confidence that the variants are present, based on an initial calculation of their likelihood give the data.
These are variants such that for any reasonable value of the site-specific prior probability of mutation, the evidence for the variant in the read data erases any doubt about the presence of the variant. 
Let $\mathrm{D}$ be the set of high confidence variant calls, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
  $$
    \alpha^{\prime}_{i,j} = \alpha_{i,j} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{i,j}\}.
  $$
Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = C_{i} \mid M,\mathrm{D})$ as
\begin{equation}
  \label{eqn:post_pred}
  $$
  p(C = C_{i} \mid M,\mathrm{D}) = \frac{\sum_{j}\alpha^{\prime}_{i,j}}{\sum_{i,j}\alpha^{\prime}_{i,j}}.
  $$
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C_{i}$, $p(m = j \mid C_{i},\mathrm{D})$ is then
\begin{equation}
  \label{eqn:to_allele}
  $$
   p(m = j \mid C_{i},\mathrm{D}) = \frac{\alpha^{\prime}_{i,j}}{\sum_{j} \alpha^{\prime}_{i,j}}.
  $$
\end{equation}
The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = C_{i})$, the prior probability of seeing a particular tri-nucleotide context.
We compute this as simply the proportion of all tri-nucleotide contexts with the context $C_{i}$ present in the genome.

>>>>

\subsection{Estimation of the mutation profile}
There are many aspects of genomic architecture which can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which has been shown to have a strong effect on the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context at a particular genomic site is made up of the identity of the reference base and the identity of the 3` and 5` flanking bases.
There are four possible bases {\{A,C,T,G\}} 3` and 5` of the site, leading to 16 possible flanking base combinations.
Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site {\{C,T\}}, resulting in $2 \cdot 16 = 32$ possible tri-nucleotide contexts.
We define a mutation context $C$ indexed by $c=\{1 \dots 32\}$, to be the tri-nucleotide context of a particular site in the genome. 
For each context, a mutation can be to any of the three possible alternate alleles {\{A,(C/T),G\}}. 
We can thus define $\mathrm{S}$, the substitution type, indexed by $c=\{1 \dots 32\}$ contexts and $m = \{1 \dots 3\}$ alternate bases, resulting in 96 possible substitution types.
Using the substitution type $\mathrm{S}$, we can rewrite equation \ref{eqn:full_prior} as

\begin{equation}
  \label{eqn:detailed_prior}
  $$
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S_{m,c}}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{p(M)}{p(C = c)},
  \end{array}
  $$
\end{equation}
where $\mathrm{p}(\mathrm{S_{m,c}})$ is the probability that a substitution of type $\mathrm{S_{m,c}}$ will occur.

@@

Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all observed tumor mutations of substitution type $\mathrm{S}_{i,j}$.
The distribution of observed substitutions $ \mathrm{p}(\mathrm{S}) $, is multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{m,c}\} $.
In a tumor with a very large number of verified mutations $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile, but in practice we find that many elements of $\boldsymbol{\pi}$ will have $0$ or a very small number of mutations present in the tumor.
As a result we need to model $\mathrm{p}(\mathrm{S})$ as dirichlet - multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$,

@@

In order to estimate $\boldsymbol{\pi}$, we generate a subset of variants in the tumor in which we have high confidence that the variants are present, based on an initial calculation of their likelihood give the data.
These are variants such that for any reasonable value of the site-specific prior probability of mutation, the evidence for the variant in the read data erases any doubt about the presence of the variant. 
Let $\mathrm{D}$ be the set of high confidence variant calls, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
  $$
    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\}.
  $$
Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = c \mid M,\mathrm{D})$ as
\begin{equation}
  \label{eqn:post_pred}
  $$
  p(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
  $$
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$, $p(m  \mid C = c,\mathrm{D})$ is then
\begin{equation}
  \label{eqn:to_allele}
  $$
   p(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
  $$
\end{equation}
The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = c)$, the prior probability of seeing a particular tri-nucleotide context.
We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the genome.
